Spike-and-Slab Generalized Additive Models and Scalable Algorithms for High-Dimensional Data

### 2021-09-23 Manuscript/01-Intro.Rmd
Revise the manuscript

Nonparametric regression models are appropriate alternatives to the "black-box" models thanks to their balance between model flexibility and interpretability. Among those, generalized additive models (GAMs), proposed in the seminal work of Hastie and Tibshirani [@Hastie1987], grew to be one of the most popular modeling tools. In a GAM, the response variable, $Y$, which is assumed to follow some exponential family distribution $EF(\mu, \phi)$ with mean $\mu$ and dispersion $\phi$, can be expressed as the summation of smoothing functions, $B_j, j = 1, \dots, p$, of a given $p$-dimensional vector of covariates $\bs x$, written as 
$$
 E(Y|\bs x) = g^{-1}(a + \sum\limits^p_{j=1}B_j(x_j)),
$$
where $g^{-1}$ is the inverse of a monotonic link function. The smoothing functions can take many forms and are estimated using a pseudo-weighted version of the backfitting algorithm [@Breiman1985]. Nevertheless, the classical GAMs cannot fulfill the increasing analytic demands for high-dimensional data analysis in biomedical studies.

@@

To address these challenges, we propose a novel Bayesian hierarchical generalized additive model (BHAM) for high dimensional data analysis. Specifically, we incorporate smoothing penalties in the model via re-parameterization of the smoothing function to avoid overly shrinking basis function coefficients. Smoothing penalties are commonly implemented in the classical GAMs through smoothing regression splines. They are quadratic norms of the coefficients and allow locally adaptive penalties on each smoothing function. A smoothing penalty conditioning on a smoothing parameter $\lambda_j$ is a function of the integration of the second derivative of the spline function, expressed mathematically as
\begin{equation}\label{eq:smoothpen}
  \text{pen}\left[f_j(x)\right] = \lambda_j \int f^{\prime\prime}_j(x)dx = \lambda_j \bs \beta_j^T \bs S_j \bs \beta_j ,
\end{equation}
where $\bs S_j$ is a known smoothing penalty matrix, and $\bs \beta_j$ are the basis function coefficients. Smoothing penalties were also previously used in the spike-and-slab GAM [@Scheipl2012] and the sparsity-smoothness penalty [@Meier2009]. Moreover, incorporating the smoothing penalty allows the separation of the linear space of smoothing functions, also know as the null space, from the nonlinear space. We then impose a new bi-part spike-and-slab spline prior on the smoothing functions for bi-level selection such that the linear and nonlinear spaces of smoothing functions can be selected separately. The prior setup encourages a flexible solution, rending one of three possibilities: no effect, only linear effect, or nonlinear effect of a predictor. In addition, two fast computing optimization-based algorithms are developed for high and ultrahigh-dimensional[@Fan2009] settings respectively.

@@

The proposed framework, BHAM, differs from previous spike-and-slab based GAMs, i.e. the spike-and-slab GAM [@Scheipl2012] and the SB-GAM [@Bai2021] in many ways. First of all, the proposed spike-and-slab spline prior is a spike-and-slab lasso type prior using independent mixture double exponential distribution, compared to spike-and-slab GAM. Spike-and-slab lasso priors provide computational convenience during model fitting by using optimization algorithms instead of computational intensive sampling algorithms. They make fitting high- and ultrahigh-dimensional models more feasible without sacrificing performance in prediction and variable selection. Secondly, the optimization algorithms introduced in this article include EM-Coordinate Descent (EM-CD) algorithm and EM-Iterative weighted least square (EM-IWLS) algorithm for different utilities. The applications of the two aforementioned algorithm are fruitful in Bayesian hierarchical models in high-dimensional data analysis. [@Yi2012; @Rockova2014a; @Rockova2018] Even though SB-GAM with a group spike-and-slab lasso prior also uses EM-CD for model fitting, the proposed EM-IWLS provides uncertainty measures that could be further used for confidence bound calculation and hypothesis testing. EM-IWLS complements the lack of inference potential from EM-CD in a median and high dimensional setting. Furthermore, the proposed model addresses the incapability of bi-level selection in SB-GAM. Lastly, the R implementation of the Bayesian hierarchical additive model is available publicly via BHAM at [https://github.com/boyiguo1/BHAM](https://github.com/boyiguo1/BHAM), make transnational science more accessible.

>>>>

Nonparametric regression models are appropriate alternatives to the "black-box" models thanks to their balance between model flexibility and interpretability. Among those, generalized additive models (GAMs), proposed in the seminal work of Hastie and Tibshirani [@Hastie1987], grew to be one of the most popular modeling tools. In a GAM, the response variable, $Y$, which is assumed to follow some exponential family distribution $EF(\mu, \phi)$ with mean $\mu$ and dispersion $\phi$, can be modeled with the summation of smoothing functions, $B_j(\cdot), j = 1, \dots, p$, of a given $p$-dimensional vector of covariates $\bs x$, written as 
$$
 E(Y|\bs x) = g^{-1}(a + \sum\limits^p_{j=1}B_j(x_j)),
$$
where $g^{-1}(\cdot)$ is the inverse of a monotonic link function. The smoothing functions can take many forms and are estimated using a pseudo-weighted version of the backfitting algorithm [@Breiman1985]. Nevertheless, the classical GAMs cannot fulfill the increasing analytic demands for high-dimensional data analysis in biomedical studies.

@@

To address these challenges, we propose a novel Bayesian hierarchical generalized additive model (BHAM) for high dimensional data analysis. Specifically, we incorporate smoothing penalties in the model via re-parameterization of the smoothing function to avoid overly shrinking basis function coefficients. Smoothing penalties are commonly implemented in the classical GAMs through smoothing regression splines. They are quadratic norms of the coefficients and allow locally adaptive penalties on each smoothing function. A smoothing penalty conditioning on a smoothing parameter $\lambda_j$ is a function of the integration of the second derivative of the spline function, expressed mathematically as
\begin{equation}\label{eq:smoothpen}
  \text{pen}\left[B_j(x)\right] = \lambda_j \int B^{\prime\prime}_j(x)dx = \lambda_j \bs \beta_j^T \bs S_j \bs \beta_j ,
\end{equation}
where $\bs S_j$ is a known smoothing penalty matrix, and $\bs \beta_j$ are the basis function coefficients. Smoothing penalties were also previously used in the spike-and-slab GAM [@Scheipl2012] and the sparsity-smoothness penalty [@Meier2009]. Moreover, incorporating the smoothing penalty allows the separation of the linear space of smoothing functions from the nonlinear space. We then impose a new bi-part spike-and-slab spline prior on the smoothing functions for bi-level selection such that the linear and nonlinear spaces of smoothing functions can be selected separately. The prior setup encourages a flexible solution, rending one of three possibilities: no effect, only linear effect, or nonlinear effect of a predictor. In addition, two scalable optimization-based algorithms are developed for high and ultrahigh-dimensional[@Fan2009] settings respectively.

@@

The proposed framework, BHAM, differs from previous spike-and-slab based GAMs, i.e. the spike-and-slab GAM [@Scheipl2012] and the SB-GAM [@Bai2021] in many ways. First of all, the proposed spike-and-slab spline prior is a spike-and-slab lasso type prior using independent mixture double exponential distribution, compared to spike-and-slab GAM. Spike-and-slab lasso priors provide computational convenience during model fitting by using optimization algorithms instead of computational intensive sampling algorithms. They make fitting high- and ultrahigh-dimensional models more feasible without sacrificing performance in prediction and variable selection. Secondly, the optimization algorithms introduced in this article include EM-Coordinate Descent (EM-CD) algorithm and EM-Iterative weighted least square (EM-IWLS) algorithm for different utilities. The applications of the two aforementioned algorithms are fruitful in Bayesian hierarchical models for high-dimensional data. [@Yi2012; @Rockova2014a; @Rockova2018] Even though SB-GAM with a group spike-and-slab lasso prior also uses EM-CD for model fitting, the proposed EM-IWLS provides uncertainty measures that could be further used for confidence bound calculation and hypothesis testing. EM-IWLS complements the lack of inference potential from EM-CD in a median and high dimensional setting. Furthermore, the proposed model addresses the incapability of bi-level selection in SB-GAM. Lastly, the R implementation of the Bayesian hierarchical additive model is available publicly via BHAM at [https://github.com/boyiguo1/BHAM](https://github.com/boyiguo1/BHAM), making transnational science more accessible.
